---
title: "Misclassification-corrected prevalence meta-analysis: depression in public safety personnel"
author: "The mcma package"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    df-print: kable
    fig-width: 7
    fig-height: 5
vignette: >
  %\VignetteIndexEntry{Case study: depression in public safety personnel}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Each sensitivity grid now fits 36 models, matching the case-study script.
# The Bayesian fits in this vignette each run 4 chains of 6,000 iterations with
# adapt_delta = 0.9999, and each sensitivity grid fits 36 models. Rendering all
# of that takes hours and needs cmdstanr/Stan, so code chunks are shown but NOT
# executed during render (eval = FALSE). Analysis chunks also use purl = FALSE
# so R CMD check skips them when running the extracted vignette script.
# The numerical results quoted in the text
# and shown in the `#>` output blocks are from the published analysis (White
# et al., 2025 data; Fawcett, Whitridge, Bartoš & Fawcett). To run it yourself,
# set eval = TRUE (and expect a long run on first build; fits are cached).
# message = FALSE keeps the Stan sampler's per-iteration progress out of the
# rendered file if you run the chunks (eval = TRUE).
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      message = FALSE)
```

## What this vignette demonstrates

Prevalence meta-analyses routinely pool studies that used brief screening
questionnaires with a smaller number that used "gold standard" diagnostic
interviews. When a disorder is rare, imperfect specificity produces false
positives among the large non-case population. These can outweigh the false
negatives caused by imperfect sensitivity, inflating the estimated prevalence
when screening and diagnostic studies are pooled naively.

`mcma` (Misclassification-Corrected Meta-Analysis) embeds a Rogan–Gladen
correction inside a Bayesian hierarchical model, propagating uncertainty in
sensitivity (*Se*) and specificity (*Sp*) into the pooled estimate. This vignette
walks through the package's recommended workflow on a real
dataset - depression among public safety personnel (police, fire, paramedic,
correctional, and communications workers) from White et al. (2025) - and
reproduces the headline result of the accompanying paper:

> A naive pooled prevalence of 12.2% falls to roughly 2.7% once
> misclassification is modelled, converging with the gold-standard diagnostic
> interviews and with general-population estimates of major depressive disorder.

The workflow:

1. **Assemble Se/Sp priors** for the screening instruments.
2. **Estimate Se/Sp from the data** where gold-standard studies are available
   (the joint model `M8`, plus a discrete kernel-weighted grid averaging (KWGA) diagnostic).
3. **Run a sensitivity analysis** across a grid of *Se*/*Sp* prior centres.
4. **Report multiple estimates** - naive, corrected, joint, and
   gold-standard-only - and compare.

## Setup

```{r, purl = FALSE}
#| label: libs
# install.packages("remotes"); remotes::install_github("jmfawcet/mcma")
library(mcma)
library(dplyr)
library(ggplot2)

# Use cmdstanr if available (recommended for the identity-link models).
options(brms.backend = "cmdstanr")

PREV_CENTER <- 0.06   # prior centre on population MDD prevalence (~6%)
```

### Data

The dataset is the depression prevalence from White et al.'s
systematic review of mental health in public safety personnel, hosted on the
project's OSF page (<https://osf.io/9y2xj/>). It has one row per prevalence
estimate, with the observed prevalence (`DepPrev`), the analytic sample size, the
instrument (`DepressionTool`), and the threshold/criterion used (`Depthreshold`).
A study is treated as gold standard when it classified cases with a
diagnostic interview (`DepressionTool2 == "diagnostic"`). The chunk below downloads it
directly from OSF (the anxiety and PTSD datasets are at
<https://osf.io/7s8b2/> and <https://osf.io/qwb25/>).

```{r, purl = FALSE}
#| label: data
# Download the depression dataset from OSF (not bundled with the package).
osf_csv <- file.path(tempdir(), "depression_alldata.csv")

if (!file.exists(osf_csv))
  download.file("https://osf.io/download/9y2xj/", osf_csv,
                mode = "wb", quiet = TRUE)

dat <- read.csv(osf_csv) %>%
  mutate(
    y     = round(DepPrev * analyticsamplesize),  # observed cases
    n     = analyticsamplesize,                    # sample size
    es_id = factor(row_number()),                  # effect-size identifier
    # Fix one mislabeled PHQ9 *before* building measure ids so it groups correctly.
    DepressionTool = recode(DepressionTool, PHQ = "PHQ9"),
    cutoff_type = case_when(
      Depthreshold == "DSMcrit"                           ~ "diagnostic",
      Depthreshold == "gpnorm"                            ~ "norm_referenced",
      Depthreshold == "notreported" | is.na(Depthreshold) ~ "unknown",
      TRUE                                                ~ "self_report_numeric"
    ),
    is_gold        = DepressionTool2 == "diagnostic",  # gold = diagnostic interview
    measure_cutoff = factor(paste(DepressionTool, Depthreshold, sep = "_"))
  ) %>%
  filter(catquality != "low")                      # drop low-quality estimates, as did White et al.

# Give every screening instrument by cutoff a measure id; gold studies share "gold".
measure_map <- dat %>%
  filter(!is_gold) %>%
  distinct(measure_cutoff) %>%
  arrange(measure_cutoff) %>%
  mutate(measure_id = as.character(row_number()))

dat <- dat %>%
  left_join(measure_map, by = "measure_cutoff") %>%
  mutate(measure_id = ifelse(is_gold, "gold", measure_id))

cat(sprintf(
  "Estimates: %d  |  Gold: %d  |  Screen: %d  |  Distinct measures: %d | Distinct measure + cutoff combinations: %d\n",
  nrow(dat), sum(dat$is_gold), sum(!dat$is_gold), n_distinct(dat$DepressionTool), n_distinct(dat$measure_cutoff)
))
```

```
#> Estimates: 89 | Gold: 10 | Screen: 79
```

Ten of the 89 estimates used diagnostic interviews; the remaining 79 used a
variety of screening instruments and cut-offs.

## Step 1 - Assemble Se/Sp priors

`mcma_priors()` builds a prior table with one row per measure. Where
measure-specific diagnostic-accuracy data exist, set each `se`/`sp` centre
directly; here - to demonstrate the realistic case where only a field-level
"typical" accuracy is known - we give every screening measure the same
weakly-informative centre (Se = Sp = 0.80) with effective sample size
`kappa = 100`. Gold-standard measures are flagged with `is_gold = TRUE` and
`NA` accuracy (the model fixes them to near-perfect classification).

```{r, purl = FALSE}
#| label: priors
screen_measures <- sort(unique(dat$measure_id[!dat$is_gold]))
all_measures    <- c(screen_measures, "gold")

base_priors <- mcma_priors(
  measure_id = all_measures,
  se      = c(rep(0.80, length(screen_measures)), NA),
  sp      = c(rep(0.80, length(screen_measures)), NA),
  kappa   = c(rep(100,  length(screen_measures)), NA),
  is_gold = c(rep(FALSE, length(screen_measures)), TRUE)
)

base_priors
```

::: callout-tip
## Be conservative on specificity for rare disorders
The single most consequential prior is *Sp*: at low prevalence the observed
case count is dominated by the false-positive rate `(1 − Sp)`. Over-estimating
*Sp* produces upward bias in corrected prevalence. When in doubt, err toward a
lower, wider *Sp* prior and lean on the Step 3 sensitivity analysis.
:::

## Step 2 - Estimate *Se*/*Sp* from the gold-standard studies

When some studies used diagnostic interviews, the diagnostic accuracy of the
screening studies can be partially identified from the data. `mcma` offers two
routes.

### Joint model (**M8**)

`mcma_fit_joint()` estimates prevalence, *Se*, and *Sp* simultaneously, collapsing
all screening studies into a single accuracy class and fixing the gold-standard
studies to near-perfect classification. We use the bounded parameterization
(*Se*, *Sp* ∈ [0.5, 1)), which keeps the Rogan–Gladen correction away from its
singularity, with wide priors so the data drive the *Se*/*Sp* estimates.

```{r, purl = FALSE}
#| label: m8
fit_m8 <- mcma_fit_joint(
  data            = dat,
  gold_column     = "is_gold",
  prev_center     = PREV_CENTER,
  se_prior_center = 0.80, sp_prior_center = 0.80,
  se_prior_sd     = 1.0,  sp_prior_sd     = 1.0,
  prev_re         = ~ (1 | es_id),
  sesp_re         = ~ (1 | es_id),
  correlated_re   = TRUE,
  bounded         = TRUE
)

m8_prev <- extract_prevalence(fit_m8)
m8_sesp <- extract_sesp(fit_m8, level = "global", bounded = TRUE)
```

```
#> M8 prevalence : ~2.7%  (95% CI roughly 1–6%)
#> M8 sensitivity: 0.85 [0.67, 0.97]   (prior-dominated; see note)
#> M8 specificity: 0.89 [0.85, 0.93]
```

The joint model recovers *Sp* sharply (~0.89) but leaves *Se* close to its prior - 
expected behaviour, because at low prevalence the data carry little information 
about *Se* (almost everyone is a true non-case). `plot_joint_sesp()` shows the 
joint posterior together with the iso-correction line of *Se*/*Sp* pairs compatible 
with the observed gold–screen gap:

```{r, purl = FALSE}
#| label: m8-plot
#| fig-cap: "Joint posterior of implied sensitivity and specificity (M8)."
plot_joint_sesp(fit_m8, show_marginals = TRUE, bounded = TRUE, n_draws = 20000)
```

### Three-level structure (studies within reports)

Several reports contribute more than one prevalence estimate. Adding a
report-level random effect (`refid`) above the per-estimate effect gives a
three-level structure; the corrected prevalence is essentially unchanged.

```{r, purl = FALSE}
#| label: m8-3level
fit_m8_tl <- mcma_fit_joint(
  data          = dat,
  gold_column   = "is_gold",
  prev_center   = PREV_CENTER,
  prev_re       = ~ (1 | es_id) + (1 | refid),
  sesp_re       = ~ (1 | es_id) + (1 | refid),
  correlated_re = TRUE,
  bounded       = TRUE
)

extract_prevalence(fit_m8_tl)
extract_sesp(fit_m8_tl, level = "global")
```

To predict latent prevalence in a new study while holding the report effect at
zero, supply `newdata` and retain only the study term in `re_formula`. This
interval includes study heterogeneity but excludes report heterogeneity and
binomial sampling noise; it differs from the pooled estimate above.

```{r, purl = FALSE}
#| label: m8-3level-prediction

new_study <- fit_m8_tl$data[1, , drop = FALSE]
new_study$es_id <- "new_study"

# Draw a new study effect; the omitted report effect is fixed at zero.
extract_prevalence(
  fit_m8_tl,
  newdata             = new_study,
  re_formula          = ~ (1 | es_id),
  allow_new_levels    = TRUE,
  sample_new_levels   = "gaussian",
  prediction_interval = TRUE
)
```

### Discrete kernel-weighted grid averaging (KWGA)

As a cross-check, `mcma_kwga()` scores a grid of *Se*/*Sp* pairs by how well a
Rogan–Gladen correction of the screening posterior reproduces the
gold standard posterior. It is fast and useful diagnostically, but does not
propagate uncertainty as fully as **M8**, so treat it as a sanity check only.

```{r, purl = FALSE}
#| label: kwga
scr_rate <- with(dat, sum(y[!is_gold]) / sum(n[!is_gold]))
fit_comp <- mcma_fit_comparison(
  data = dat, gold_column = "is_gold",
  gold_prev_center = PREV_CENTER, scr_prev_center = scr_rate,
  prev_re = ~ (1 | es_id)
)

kwga      <- mcma_kwga(fit_comp,
                       se_grid = seq(0.70, 1, by = 0.025),
                       sp_grid = seq(0.70, 1, by = 0.025))
kwga_prev <- mcma_kwga_prevalence(kwga, fit_comparison = fit_comp, n_draws = 4000)
kwga
```

```
#> Discrete KWGA: Sp ≈ 0.875 (Se weakly identified) - consistent with M8.
```

## Step 3 - Sensitivity analysis

Because the corrected estimate depends on the assumed *Se*/*Sp*, the workflow's core
safeguard is a prior sensitivity analysis: re-fit the corrected model across
a grid of *Se*/*Sp* prior centres and visualise how the pooled prevalence moves.
`mcma_sensitivity()` fits one model per grid cell (cached to `model_dir`).

```{r, purl = FALSE}
#| label: sens
se_vals <- seq(0.70, 0.95, by = 0.05)
sp_vals <- seq(0.70, 0.95, by = 0.05)

sens <- mcma_sensitivity(
  data        = dat,
  priors      = base_priors,
  se_grid     = se_vals,
  sp_grid     = sp_vals,
  prev_center = PREV_CENTER,
  bounded     = TRUE,
  prev_re     = ~ (1 | es_id),
  sesp_re     = ~ (1 | es_id),
  model_dir   = "models/sensitivity_case"
)

plot(sens, type = "heatmap", metric = "prev_mean")
```

A companion grid, `mcma_sensitivity_comparison()`, fits a gold vs screen model
at each cell so you can read off the difference between the corrected
screening estimate and the gold-standard estimate. Pass the sampler settings
explicitly so the grid uses the same rigorous control as the other models:

```{r, purl = FALSE}
#| label: sens-comp
sens_comp <- mcma_sensitivity_comparison(
  data        = dat,
  priors      = base_priors,
  se_grid     = se_vals,
  sp_grid     = sp_vals,
  prev_center = PREV_CENTER,
  bounded     = TRUE,
  prev_re     = ~ (1 | es_id),
  sesp_re     = ~ (1 | es_id),
  gold_column = "is_gold",
  model_dir   = "models/sensitivity_comp_case",
  chains = 4, cores = 4, iter = 6000, warmup = 3000,
  adapt_delta = 0.9999, max_treedepth = 20
)

plot(sens_comp, type = "difference", include_cell_estimates = TRUE)
```

Across the plausible region (*Se*/*Sp* roughly 0.70–0.90) the corrected prevalence
is stable and low, and the corrected screening estimate is statistically
indistinguishable from the gold-standard estimate. The estimate only climbs once
*Sp* is pushed above ~0.90 - higher than is typical for depression screeners at
standard cut-offs.

## Step 4 - Report multiple estimates

Finally, report and compare the full set: the naive benchmark (**M1**), the
gold-standard-only anchor, the joint model (**M8**), the corrected model
with priors centred at the **M8** estimates (**M6**), and the KWGA diagnostic.

```{r, purl = FALSE}
#| label: report-fits
fit_naive <- mcma_fit(dat, priors = NULL, prev_center = PREV_CENTER,
                      prev_re = ~ (1 | es_id))

fit_gold  <- mcma_fit(filter(dat, is_gold), priors = NULL,
                      prev_center = PREV_CENTER, prev_re = ~ (1 | es_id))

# Centre the corrected model's Se/Sp priors at the M8-implied values.
m6_priors <- update(base_priors, se = m8_sesp$se_mean, sp = m8_sesp$sp_mean)

fit_m6 <- mcma_fit(
  data = dat, priors = m6_priors, prev_center = PREV_CENTER,
  prev_re = ~ (1 | es_id), sesp_re = ~ (1 | es_id),
  correlated_re = TRUE, bounded = TRUE
)
```

```{r, purl = FALSE}
#| label: report-table
prev_raw <- bind_rows(
  extract_prevalence(fit_naive) %>% mutate(model = "M1 (Naive)"),
  extract_prevalence(fit_gold)  %>% mutate(model = "Gold-only"),
  extract_prevalence(fit_m8)    %>% mutate(model = "M8 (Joint)"),
  extract_prevalence(fit_m6)    %>% mutate(model = "M6 (M8-informed)"),
  tibble(model = "KWGA (analytic)",
         mean = kwga_prev$summary$mean,  median = kwga_prev$summary$median,
         l95  = kwga_prev$summary$ci_lb, u95    = kwga_prev$summary$ci_ub)
)

prev_table <- prev_raw %>%
  transmute(model,
            `Prevalence (%)` = sprintf("%.1f", mean * 100),
            `95% CI` = sprintf("[%.1f, %.1f]", l95 * 100, u95 * 100))
prev_table
```

```
#> model              Prevalence (%)
#> M1 (Naive)         12.2
#> Gold-only           2.5
#> M8 (Joint)          2.7
#> M6 (M8-informed)    2.2
#> KWGA (analytic)     1.6
#>
#> The 95% CIs of the corrected models overlap one another and the gold-only
#> anchor in every case, and none of them overlaps the naive estimate.
```

The corrected point estimates range from approximately 1.6% to 2.7%, compared
with the naive 12.2%. Their 95% intervals overlap one another and the
gold-standard-only anchor, and include values within the published
general-population diagnostic interview range for MDD (≈3.1–5.4%). The KWGA
distribution has a median of zero and a 95% interval of approximately 0–8%.
A forest plot makes the gap between the naive estimate and everything else clear:

```{r, purl = FALSE}
#| label: forest
#| fig-cap: "Pooled depression prevalence by model, with 95% intervals."
model_order <- c("M1 (Naive)", "Gold-only", "KWGA (analytic)",
                 "M8 (Joint)", "M6 (M8-informed)")

prev_raw %>%
  mutate(model = factor(model, levels = rev(model_order))) %>%
  ggplot(aes(x = mean * 100, y = model)) +
  geom_errorbarh(aes(xmin = l95 * 100, xmax = u95 * 100), height = 0.18) +
  geom_point(size = 2.6) +
  labs(x = "Prevalence (%)", y = NULL) +
  theme_minimal(base_size = 12)
```


## Diagnostics

Always check standard MCMC diagnostics and posterior predictive fit alongside the
substantive output.

```{r, purl = FALSE}
#| label: diagnostics
mcma_convergence(M1 = fit_naive, Gold = fit_gold, Comparison = fit_comp,
                 M8 = fit_m8, M8_3level = fit_m8_tl, M6 = fit_m6)

# Posterior predictive checks: mean discrepancy ≈ 0, variance ratio ≈ 1,
# Bayesian p-values ≈ 0.5 indicate adequate fit.
bind_rows(
  mcma_ppc(fit_naive) %>% mutate(model = "M1 (Naive)"),
  mcma_ppc(fit_gold)  %>% mutate(model = "Gold-only"),
  mcma_ppc(fit_m8)    %>% mutate(model = "M8 (Joint)"),
  mcma_ppc(fit_m6)    %>% mutate(model = "M6 (M8-informed)")
)
```

## References

- Fawcett, Whitridge, Bartoš & Fawcett. *Bayesian meta-analysis of disorder
  prevalence with misclassification correction.*
- White et al. (2025). Depression in public safety personnel (case-study data).
- Rogan & Gladen (1978). Estimating prevalence from the results of a screening test.

```{r}
#| label: session
#| eval: true
sessionInfo()
```
